Complex Synapse Identification

### 2014-07-03 Notes/Identification.tex
notes on sparse priors

When we don't have any prior information about the model $\M$, we fit the model by maximising the likelihood function:
%
\begin{equation}\label{eq:like}
  \pr\cond{\w^A[0,T]}{\M} = \prod_a \pr\cond{\w^a[0,T]}{\M}.
\end{equation}
%
To do this, we define an axillary function
%
\begin{equation}\label{eq:emq}
  Q(\M,\M') = \sum_{i^A[0,T]} \pr\cond{i^A[0,T],\w^A[0,T]}{\M} \log\!\! \brk{
      \frac{ \pr\cond{i^A[0,T],\w^A[0,T]}{\M'} }{ \pr\cond{i^A[0,T],\w^A[0,T]}{\M} } }.
\end{equation}
%
It can be shown that, if $\M'$ is chosen to maximise this quantity (which is easier than maximising the likelihood), the it will have a higher likelihood than $\M'$.
In effect, $\M$ is used to estimate the hidden state trajectories, which are then used to re-estimate the model.
This can be done iteratively to find a local maximum of the likelihood.
If this is repeated several times with random initialisation, we can hope to find the global maximum.

@@

These quantities usually suffer from underflow.
This can be avoided by using normalised forwar/backward variables:
%
\begin{equation}\label{eq:normalphabeta}
\begin{aligned}
  \tilde{\alpha}_i(t) &= \alpha_i(t) \prod_{s=0}^{t} \eta(s), \\
  \tilde{\beta}_i(t) &= \beta_i(t) \prod_{s=t+1}^{T} \eta(s),
\end{aligned}
\end{equation}
%
where $\eta[0,T]$ is chosen so that all of the $\tilde{\alpha}[0,T]$ are normalised: $\sum_i \tilde{\alpha}_i(t) = 1$.
These quantities have the interpretations:
%
\begin{equation}\label{eq:normintepr}
\begin{aligned}
  \tilde{\alpha}_{i(t)}(t) &= \pr\cond{i(t)}{\w[0,t],\M}, \\
  \eta(t) &= \pr\cond{\w(t)}{\w[0,t-1],\M}, \\
  \prod_{s=0}^t \eta(t) &= \pr\cond{\w[0,t]}{\M}.
\end{aligned}
\end{equation}
%
Note that the new backward variables $\tilde{\beta}$ are not normalised and have no particularly interesting interpretation.
However:
%
\begin{equation}\label{eq:normstateprob}
\begin{aligned}
  \widetilde{N}^\nu_{i(t)i(t+1)} &= \sum_{t=0}^{T-1} \pr\cond{i[t,t+1])}{\w[0,T],\M} \\
     &= \sum_{t=0}^{T-1} \delta_{\mu(t)\nu} \, \tilde{\alpha}_i(t) \, \tilde{\beta}_j(t+1) \, \eta(t+1) \, \M^{\nu}_{i(t)i(t+1)} \, \Pib(t+1) ,\\
  \pr\cond{i(t)}{\w[0,T],\M} &= \alpha_{i(t)}(t) \, \beta_{i(t)}(t).
\end{aligned}
\end{equation}
%
Then we can write
%
\begin{equation}\label{eq:singlediffsnorm}
\begin{aligned}
  \pdiff{\CL}{\M'^\mu_{ij}} &= \pr\cond{\w[0,T]}{\M} \, \frac{ \widetilde{N}^\mu_{ij} }{ \M'^\mu_{ij} } - \lambda^\mu_{i}, \\
  \pdiff{\CL}{\pib'_{i}} &= \pr\cond{\w[0,T]}{\M} \, \frac{ \tilde{\alpha}_i(0) \, \tilde{\beta}_i(0) }{ \pib'_{i} } - \lambda,
\end{aligned}
\end{equation}
%
The factors of $\pr\cond{\w[0,T]}{\M}$ can be absorbed by a redefinition of the Lagrange multipliers.

Demanding a maximum and setting the derivatives to zero leaves us with the update rule:
%
\begin{equation}\label{eq:BWupdate}
\begin{aligned}
  \M'^\mu_{ij} &= \frac{ \widetilde{N}^\mu_{ij} }{ \sum_k \widetilde{N}^\mu_{ik} },\\
  \pib'_{i} &= \frac{ \tilde{\alpha}_i(0) \, \tilde{\beta}_i(0) }{ \sum_j \tilde{\alpha}_j(0) \, \tilde{\beta}_j(0) }.
\end{aligned}
\end{equation}
%
This can be iterated until a maximum is found.

@@

Defining $\widetilde{N}^{A\mu}_{ij} = \sum_a \widetilde{N}^{a\mu}_{ij}$ and setting the derivatives to zero leaves us with the update rule:
%
\begin{equation}\label{eq:RJupdate}
\begin{aligned}
  \M'^\mu_{ij} &= \frac{ \widetilde{N}^{A\mu}_{ij} }{ \sum_k \widetilde{N}^{A\mu}_{ik} },\\
  \pib'_{i} &= \frac{ \sum_a \tilde{\alpha}^a_i(0) \, \tilde{\beta}^a_i(0) }{ \sum_{bj} \tilde{\alpha}^b_j(0) \, \tilde{\beta}^b_j(0) }.
\end{aligned}
\end{equation}
%
This can be iterated until a maximum is found.

@@

We will consider priors of the form
%
\begin{equation}\label{eq:priorsum}
  \pr\!\prn{\M} = \frac{\exp\prn{-\beta\sum_{\mu ij}E^\mu_{ij}(\M^\mu_{ij})}}{Z(\beta)}.
\end{equation}
%
We will not put any prior on $\pib$, so its update rule will be unaffected.
The normalisation constant $Z(\beta)$ will not play any role either.
Now the derivatives read:
%
\begin{equation}\label{eq:multidiffs}
  \pdiff{\CL}{\M'^\mu_{ij}} 
    = \pr\!\prn{\w[0,T],\M} \prn{ \frac{ \widetilde{N}^{A\mu}_{ij} }{ \M'^\mu_{ij} }  - \pdiff{E^\mu_{ij}}{\M'^\mu_{ij}} }- \lambda^\mu_{i}.
\end{equation}
%
Once again, we can absorb the factors of $\pr\!\prn{\w[0,T],\M}$ by a redefinition of the Lagrange multipliers.

>>>>

When we don't have any prior information about the model $\M$, we fit the model by maximising the likelihood function:
%
\begin{equation}\label{eq:like}
  \pr\cond{\w^A[0,T]}{\M} = \prod_a \pr\cond{\w^a[0,T]}{\M}.
\end{equation}
%
To do this, we define an axillary function \cite{Baum1970baumwelch,Dempster2007EM}:
%
\begin{equation}\label{eq:emq}
  Q(\M,\M') = \sum_{i^A[0,T]} \pr\cond{i^A[0,T],\w^A[0,T]}{\M} \log\!\! \brk{
      \frac{ \pr\cond{i^A[0,T],\w^A[0,T]}{\M'} }{ \pr\cond{i^A[0,T],\w^A[0,T]}{\M} } }.
\end{equation}
%
It can be shown that, if $\M'$ is chosen to maximise this quantity (which is easier than maximising the likelihood), the it will have a higher likelihood than $\M'$.
In effect, $\M$ is used to estimate the hidden state trajectories, which are then used to re-estimate the model.
This can be done iteratively to find a local maximum of the likelihood.
If this is repeated several times with random initialisation, we can hope to find the global maximum.

@@

These quantities usually suffer from underflow.
This can be avoided by using normalised forwar/backward variables \cite{Zhai2003HMMnorm}:
%
\begin{equation}\label{eq:normalphabeta}
\begin{aligned}
  \tilde{\alpha}_i(t) &= \alpha_i(t) \prod_{s=0}^{t} \eta(s), \\
  \tilde{\beta}_i(t) &= \beta_i(t) \prod_{s=t+1}^{T} \eta(s),
\end{aligned}
\end{equation}
%
where $\eta[0,T]$ is chosen so that all of the $\tilde{\alpha}[0,T]$ are normalised: $\sum_i \tilde{\alpha}_i(t) = 1$.
These quantities have the interpretations:
%
\begin{equation}\label{eq:normintepr}
\begin{aligned}
  \tilde{\alpha}_{i(t)}(t) &= \pr\cond{i(t)}{\w[0,t],\M}, \\
  \eta(t) &= \pr\cond{\w(t)}{\w[0,t-1],\M}, \\
  \prod_{s=0}^t \eta(t) &= \pr\cond{\w[0,t]}{\M}.
\end{aligned}
\end{equation}
%
Note that the new backward variables $\tilde{\beta}$ are not normalised and have no particularly interesting interpretation.
However:
%
\begin{equation}\label{eq:normstateprob}
\begin{aligned}
  \widetilde{N}^\nu_{i(t)i(t+1)} &= \sum_{t=0}^{T-1} \pr\cond{i[t,t+1])}{\w[0,T],\M} \\
     &= \sum_{t=0}^{T-1} \delta_{\mu(t)\nu} \, \tilde{\alpha}_i(t) \, \tilde{\beta}_j(t+1) \, \eta(t+1) \, \M^{\nu}_{i(t)i(t+1)} \, \Pib(t+1) ,\\
  \pr\cond{i(t)}{\w[0,T],\M} &= \alpha_{i(t)}(t) \, \beta_{i(t)}(t).
\end{aligned}
\end{equation}
%
Then we can write
%
\begin{equation}\label{eq:singlediffsnorm}
\begin{aligned}
  \pdiff{\CL}{\M'^\mu_{ij}} &= \pr\cond{\w[0,T]}{\M} \, \frac{ \widetilde{N}^\mu_{ij} }{ \M'^\mu_{ij} } - \lambda^\mu_{i}, \\
  \pdiff{\CL}{\pib'_{i}} &= \pr\cond{\w[0,T]}{\M} \, \frac{ \tilde{\alpha}_i(0) \, \tilde{\beta}_i(0) }{ \pib'_{i} } - \lambda,
\end{aligned}
\end{equation}
%
The factors of $\pr\cond{\w[0,T]}{\M}$ can be absorbed by a redefinition of the Lagrange multipliers.

Demanding a maximum and setting the derivatives to zero leaves us with the update rule \cite{Baum1970baumwelch}:
%
\begin{equation}\label{eq:BWupdate}
\begin{aligned}
  \M'^\mu_{ij} &= \frac{ \widetilde{N}^\mu_{ij} }{ \sum_k \widetilde{N}^\mu_{ik} },\\
  \pib'_{i} &= \frac{ \tilde{\alpha}_i(0) \, \tilde{\beta}_i(0) }{ \sum_j \tilde{\alpha}_j(0) \, \tilde{\beta}_j(0) }.
\end{aligned}
\end{equation}
%
This can be iterated until a maximum is found.

@@

Defining $\widetilde{N}^{A\mu}_{ij} = \sum_a \widetilde{N}^{a\mu}_{ij}$ and setting the derivatives to zero leaves us with the update rule \cite{rabiner1993speechrec}:
%
\begin{equation}\label{eq:RJupdate}
\begin{aligned}
  \M'^\mu_{ij} &= \frac{ \widetilde{N}^{A\mu}_{ij} }{ \sum_k \widetilde{N}^{A\mu}_{ik} },\\
  \pib'_{i} &= \frac{ \sum_a \tilde{\alpha}^a_i(0) \, \tilde{\beta}^a_i(0) }{ \sum_{bj} \tilde{\alpha}^b_j(0) \, \tilde{\beta}^b_j(0) }.
\end{aligned}
\end{equation}
%
This can be iterated until a maximum is found.

@@

We will consider priors of the form
%
\begin{equation}\label{eq:priorsum}
  \pr\!\prn{\M} = \frac{\exp\prn{-\beta\sum_{\mu ij}E^\mu_{ij}(\M^\mu_{ij})}}{Z(\beta)}.
\end{equation}
%
We will not put any prior on $\pib$, so its update rule will be unaffected.
The normalisation constant $Z(\beta)$ will not play any role either.
Now the derivatives read:
%
\begin{equation}\label{eq:priordiffs}
  \pdiff{\CL}{\M'^\mu_{ij}} 
    = \pr\!\prn{\w[0,T],\M} \prn{ \frac{ \widetilde{N}^{A\mu}_{ij} }{ \M'^\mu_{ij} }  - \beta \pdiff{E^\mu_{ij}}{\M'^\mu_{ij}} }- \lambda^\mu_{i}.
\end{equation}
%
Once again, we can absorb the factors of $\pr\!\prn{\w[0,T],\M}$ by a redefinition of the Lagrange multipliers.

\subsection{Off-diagonal $\CL^1$ penalty \label{sec:offdiagL1}}

Here we use the off-diagonal $\CL^1$ norm as a penalty:
%
\begin{equation}\label{eq:L1penalty}
  E^\mu_{ij} = (1-\delta_{ij}) \M^\mu_{ij}.
\end{equation}
%
Then the condition for a maximum \eqref{eq:priordiffs} reads
%
\begin{equation}\label{eq:L1priormax}
\begin{aligned}
  \frac{ \widetilde{N}^{A\mu}_{ij} }{ \M'^\mu_{ij} }  - (1-\delta_{ij})\beta - \lambda^\mu_{i} &= 0, \\
   \M'^\mu_{ij} &= \frac{\widetilde{N}^{A\mu}_{ij}}{\lambda^\mu_{i}+(1-\delta_{ij})\beta}.
\end{aligned}
\end{equation}
%
Demanding normalisation gives
%
\begin{multline}\label{eq:L1norm}
  \sum_{j \neq i}\frac{\widetilde{N}^{A\mu}_{ij}}{\lambda^\mu_{i}+\beta} +\frac{\widetilde{N}^{A\mu}_{ii}}{\lambda^\mu_{i}} =0, \\
  \implies \quad
  \lambda^\mu_{i} = \half \brk{ \prn{\sum_j \widetilde{N}^{A\mu}_{ij} -\beta} + 
     \sqrt{\prn{\sum_j \widetilde{N}^{A\mu}_{ij} -\beta}^2 +4\beta\widetilde{N}^{A\mu}_{ii}}}
\end{multline}
%
Which produces the update rule
%
\begin{equation}\label{eq:L1update}
\begin{aligned}
  \M'^\mu_{ij} &= \frac{ 2\widetilde{N}^{A\mu}_{ij} }{ \prn{\sum_k \widetilde{N}^{A\mu}_{ik} +\beta} +
     \sqrt{\prn{\sum_k \widetilde{N}^{A\mu}_{ik} -\beta}^2 +4\beta\widetilde{N}^{A\mu}_{ii}} } 
     \qquad \text{for } i\neq j, \\
  \M'^\mu_{ii} &= \frac{ 2\widetilde{N}^{A\mu}_{ii} }{ \prn{\sum_k \widetilde{N}^{A\mu}_{ik} -\beta} +
     \sqrt{\prn{\sum_k \widetilde{N}^{A\mu}_{ik} -\beta}^2 +4\beta\widetilde{N}^{A\mu}_{ii}} }.
\end{aligned}
\end{equation}
%

\subsection{$\CL^\half$ penalty \label{sec:Lhalf}}

Here we use the  $\CL^\half$ norm as a penalty:
%
\begin{equation}\label{eq:Lhalfpenalty}
  E^\mu_{ij} = 2\sqrt{\M^\mu_{ij}}.
\end{equation}
%
Then the condition for a maximum \eqref{eq:priordiffs} reads
%
\begin{equation}\label{eq:Lhalfpriormax}
  \frac{ \widetilde{N}^{A\mu}_{ij} }{ \M'^\mu_{ij} }  - \frac{\beta}{\sqrt{\M'^\mu_{ij}}} - \lambda^\mu_{i} = 0.
\end{equation}
%
It is helpful to define
%
\begin{equation}\label{eq:Lhalfvars}
\begin{aligned}
  \widetilde{N}^{A\mu}_{ij} &= \beta^2 L^\mu_{ij}, \\
  \lambda^\mu_i &= \prn{\gamma^\mu_i}^{-2}.
\end{aligned}
\end{equation}
%
Then we have the update rule
%
\begin{equation}\label{eq:Lhalfupdate}
  \M'^\mu_{ij} = \frac{ \prn{\gamma^\mu_i}^2 \beta^2 }{ 4 } \prn{\sqrt{ \prn{\gamma^\mu_i}^2 + 4\prn{L^\mu_{ij}}^2 } - \gamma^\mu_i }^2, 
\end{equation}
%
with $\gamma^\mu_i$ determined by:
%
\begin{equation}\label{eq:Lhalfnorm}
    \frac{ \prn{\gamma^\mu_i}^2 \beta^2 }{ 4 } \sum_j \prn{\sqrt{\prn{\gamma^\mu_i}^2 + 4\prn{L^\mu_{ij}}^2} - \gamma^\mu_i}^2 = 1.
\end{equation}
%
